```{r head, child="metadata.yaml"}
```

```{r setup, echo = FALSE}
require(knitr)
options(width=60, width.cutoff=60)
opts_chunk$set(tidy=TRUE)

hook_source_def = knit_hooks$get('source')
knit_hooks$set(source = function(x, options){
  if (!is.null(options$verbatim) && options$verbatim){
    opts = gsub(",\\s*verbatim\\s*=\\s*TRUE\\s*", "", options$params.src)
    bef = sprintf('\n\n    ```{r %s}\n', opts, "\n")
    stringr::str_c(bef, paste(knitr:::indent_block(x, "    "), collapse = '\n'), "\n    ```\n")
  } else {
     hook_source_def(x, options)
  }
})

rinline <- function(code){
	sprintf('`r %s`', code)
}
```

```{r numbered_chunk_hook, eval=FALSE, echo=FALSE}

# The current chunk hook may not be the default, and should be processed
# prior to doing the numbering; which should come last.

previous_chunk_hook <- knitr::knit_hooks$get("chunk")

knitr::knit_hooks$set(chunk = function(x, options) {
    
    x <- previous_chunk_hook(x, options)
    
    if (isTRUE(options$number)) {
        
        str <- "{.r .numberLines"
        if (!is.null(options$startFrom)) {
            str <- paste0(str, " startFrom=\"", options$startFrom, "\"")
        }
        str <- paste0(str, "}")
        
        x <- gsub("(\\s?[`]{3,})r", paste0("\\1", str), x)
        
    }
    
    return(gsub("(^\n|\n+$)", "", x))
    
})
```
````{r numbered_chunk_hook, echo=FALSE}
````

# Introduction
This is an introduction to the use of [Markdown](http://daringfireball.net/projects/markdown/)
with embedded R code to create dynamic documents in multiple formats, 
e.g. HTML, PDF and Word. This is useful to generate reports (or papers) that contain all
the relevant R code to carry out the analysis and allows for automatic updates to the 
document if either the code or the data change. As a result analyses become a lot 
easier to reproduce because the code and the presentation of results are closely 
linked and figures and tables can be updated automatically.

Several tools are available to do this but here we will mainly focus on a combination
of two of these.

[`knitr`](http://yihui.name/knitr/)
:   This R package provides functions that allow the processing of Markdown documents
    with embedded R code. The code will be executed and its output, including plots,
    can be included in the output.
[`pandoc`](http://johnmacfarlane.net/pandoc/)
:   This is a versatile tool that supports conversion between a large number of input
    and output formats. We will use 
    [pandoc's own Markdown flavour](http://johnmacfarlane.net/pandoc/demo/example9/pandocs-markdown.html)
    (which includes numerous extension to the original format) as input and use its 
    conversion capabilities to obtain output in several commonly used formats.  

Throughout this document examples of R code and Markdown formatting will be presented in 
code blocks:
```r
message("This is R code")
```

To better show the effect of Markdown examples on the output these will often be
followed by the same text rendered in the output format. To distinguish these examples
from the main text the entire block of raw and converted Markdown will be framed by
horizontal lines. 

----

```markdown
This is Markdown text in **bold** and *italics*.
```

This is Markdown text in **bold** and *italics*.

----

In addition to the code examples provided throughout this document the document itself is
written in Markdown with embedded R code and may illustrate additional features.

# Brief Markdown primer
A Markdown formatted file is in essence a plain text file that may contain a number of
formatting marks. It is designed to be easy to write and read in its raw form. Although
it was originally designed as an easier way to write web pages it can be converted to
many other rich text formats.  

The purpose of this section is to briefly describe basic elements of Markdown formatting.
More detailed descriptions are available online, e.g. at the official
[Markdown](http://daringfireball.net/projects/markdown/syntax) and
[`pandoc`](http://johnmacfarlane.net/pandoc/README.html) websites.

## Headers, paragraphs and emphasis
The basics of text formatting involve marking of text as headings, structuring 
it into paragraphs and highlighting selected words for emphasis. Headings can be
created by underlining them:

```markdown
This is a top level heading
===========================
This is some ordinary text.

This is a second level heading
------------------------------
It is followed by more ordinary text.

### Third level heading
Adding more "#" creates lower level headings
```

Note that `pandoc` also allows the use of "#" and "##" for first and second level headings. 

Paragraphs are created by adding an empty line between two lines of text:

----

```markdown
This is the first paragraph.
Line breaks are generally ignored in formatting.

This is the second paragraph.
If you add two or more spaces to the end of a line  
the line break will be preserved in the conversion
to the output format.  
```

This is the first paragraph.
Line breaks are generally ignored in formatting.

This is the second paragraph.
If you add two or more spaces to the end of a line  
the line break will be preserved in the conversion
to the output format.

----

Several methods of highlighting text are supported:

----

```markdown
Words within a paragraph and be *emphasised*. These are usually rendered in *italics*.
**Strong emphasis** typically results in **bold** text. Instead of * it is also possible
to use _ for emphasis. With pandoc it is also possible to ~~strike out~~ text.
```

Words within a paragraph and be *emphasised*. These are usually rendered in *italics*.
**Strong emphasis** typically results in **bold** text. Instead of * it is also possible
to use _ for emphasis. With pandoc it is also possible to ~~strike out~~ text.

----

## Block elements

### Block quotes
In Markdown quotes can be marked using the same conventions commonly used in email:

----

```markdown
 > This text is quoted. A single ">" at the beginning
 > of the paragraph is sufficient for the entire paragraph 
 > to be quoted (but syntax highlighting may not work properly).
 > 
 > > You can also quote other quotes, i.e. block quotes can be nested.
```
 
 > This text is quoted. A single ">" at the beginning
 > of the paragraph is sufficient for the entire paragraph 
 > to be quoted (but syntax highlighting may not work properly).
 > 
 > > You can also quote other quotes, i.e. block quotes can be nested.

----

### Lists
Basic bullet lists can be created by starting a line with a \*:

----

```markdown
 * first item
 * second item
 * third item
```
 
 * first item
 * second item
 * third item

----

Ordered lists start with numbers

----

```markdown
1. first item
2. second item
3. third item
```

1. first item
2. second item
3. third item

----

but `pandoc` also allows this:

----

```markdown
#. first item
#. second item
#. third item
```

#. first item
#. second item
#. third item

----


There is support for other of list types and variations of the basic syntax
in `pandoc`. See the [documentation](http://johnmacfarlane.net/pandoc/README.html#lists)
for more details. 

### Tables
Basic Markdown tables are created by lining up the columns and making headers, like so:

----

```markdown
 Column 1    Column 2    Column 3
 --------    --------    --------
   1            10         100
   2            20         200
   3            30         300
  
 Table: A simple table 
```


  Column 1      Column 2      Column 3
 ----------    ----------    ----------
   1              10            100
   2              20            200
   3              30            300
  
 Table: A simple table 

----

Just as with lists there are several variations and extensions to this basic syntax
supported by `pandoc`. As usual, details can be found in the
[documentation](http://johnmacfarlane.net/pandoc/README.html#tables).

### Code blocks and inline code
Special blocks to display source code with syntax highlighting can be included by starting
a line with three back ticks, optionally followed by attributes to control aspects of
the highlighting. A block like the one below will be rendered as R code.

----

````
```r
x <- seq(-6,6, by=0.1)
yNorm <- dnorm(x)
yt <- dt(x, df=3)
yCauchy <- dcauchy(x)
plot(x, yNorm, type="l", ylab="Density")
lines(x, yt, col=2)
lines(x, yCauchy, col=4)
legend("topright", legend=c("standard normal", "t (df=2)", "Cauchy"), 
    col=c(1,2,4), lty=1)
```
````

```r
x <- seq(-6,6, by=0.1)
yNorm <- dnorm(x)
yt <- dt(x, df=3)
yCauchy <- dcauchy(x)
plot(x, yNorm, type="l", ylab="Density")
lines(x, yt, col=2)
lines(x, yCauchy, col=4)
legend("topright", legend=c("standard normal", "t (df=2)", "Cauchy"), 
    col=c(1,2,4), lty=1)
```

----

Code fragments can also be included inline:

----

```markdown
This is normal text with some R code: `x <- runif(100)`{.r}.
```
This is normal text with some R code: `x <- runif(100)`{.r}.

----

# Using `knitr` for dynamic code blocks
The code blocks we have seen so far are all static, i.e. while they do include
valid source code this code is not interpreted, just displayed. To achieve the aim of
a dynamic document that can be updated automatically if the underlying data or analysis
change we need code blocks that are actually executed. The `knitr` R package does just 
that. Lets look again at the R code from the example in the previous section but this 
time we will use a code block that `knitr` will process. 

````
```{r distributions, verbatim = TRUE}
x <- seq(-6,6, by=0.1)
yNorm <- dnorm(x)
yt <- dt(x, df=3)
yCauchy <- dcauchy(x)
```
````

In this example syntac highlighting for the R code has been switched off
to better demonstrate how the code chunks are created. Once the code in the above
block has been evaluated we can use it for inline R statements that will be
replaced with the computed values. For example we can do something like this:

----

```markdown
The Normal density was evaluated at `r rinline("length(yNorm)")` points.
```

The Normal density was evaluated at `r length(yNorm)` points.

----

## Figures
To add a figure with a plot of the data all that is needed is to create the
plot in an R chunk.

```{r distribution_plot, fig.cap="Three related distributions"}
plot(x, yNorm, type="l", ylab="Density")
lines(x, yt, col=2)
lines(x, yCauchy, col=4)
legend("topright", legend=c("standard normal", "t (df=3)", "Cauchy"), 
   col=c(1,2,4), lty=1)
```

This automatically includes the plot that was generated as a figure in the
final document. It is possible to include a custom caption using the chunk option
`fig.cap`.

## Animations
It is possible to include animations (generated from a series of plots) instead of 
a single figure. Look at this slightly more complex version of the previous example:

```{r distributions2, fig.show="animate", fig.cap="Movie of three related distributions"}
x <- seq(-6,6, by=0.1)
yNorm <- dnorm(x)
yCauchy <- dcauchy(x)
par(bg="white")
for(i in 1:20){
    plot(x, yNorm, type="l", ylab="Density")
    lines(x, dt(x, df=i), col=2)
    lines(x, yCauchy, col=4)
    legend("topright", legend=c("standard normal", paste0("t (df = ", i, ")"), "Cauchy"), 
       col=c(1,2,4), lty=1)
}
```

It is possible to generate an animated GIF from this sequence of plots by wrapping the
above code in a function and then calling `saveGIF` from the animation package.

```{r distributions3, results="hide"}
threeDists <- function(df, x=seq(-6,6, by=0.1)){
    yNorm <- dnorm(x)
    yCauchy <- dcauchy(x)
    yt <- dt(x, df=df)
    par(bg="white")
    plot(x, yNorm, type="l", ylab="Density")
    lines(x, yt, col=2)
    lines(x, yCauchy, col=4)
    legend("topright", legend=c("standard normal", paste0("t (df = ", df, ")"), "Cauchy"), 
       col=c(1,2,4), lty=1)
}

plotFun <- function(df){
    threeDists(df)
    animation::ani.pause()
}

animation::saveGIF(lapply(1:20, plotFun), interval=0.5, movie.name="dist3.gif")
```

Since the graphics output of this code is written directly to a file rather than an on-screen
graphics device it will not outomatically included in the Markdown document produced 
by `knitr`. It can be included manually using the Markdown syntax for the inclusion
of figures.

```markdown
![Animated GIF of three related distributions](dist3) 
```
![Animated GIF of three related distributions](dist3)

Note that this only works for output document formats that support GIFs. As a fallback
we generate a png of the first frame to be included in other formats, e.g. PDF, that
can't display GIFs.

```{r dist3-png, results="hide"}
png("dist3.png")
threeDists(1)
dev.off()
```

Here we make use of `pandoc`'s `--default-image-extension` option to set the daefault
image format to gif for HTML and docx output and to png for PDF.

## Tables
It is often convenient to display the contents of R objects in the final document. 
While it is easy to simply display the output of R's `print` statement as it would
be displayed in the R console, this is not exactly producing pretty results. It
is much more elegant to include proper tables that can be rendered nicely in
the final document. Manually  formatting the output as a Markdown table (that pandoc
will then convert to the final output format) can be a daunting task. Fortunately
R functions exist to help with this task. The `knitr` package provides a simple function, 
`kable`, that allows automatic formatting of tables. This requires the data to be in a 
suitable format (either a `data.frame` or a `matrix`) so some preprocessing may be necessary.

Consider the [iris](http://stat.ethz.ch/R-manual/R-patched/library/datasets/html/iris.html)
dataset distributed with R.

```{r iris1, results="asis"}
data(iris)
knitr::kable(head(iris))
```

A table of summary statistics can be obtained with a little extra effort:
```{r iris2}
irisSummary <- apply(iris[,1:4], 2, function(x) tapply(x, iris$Species, summary))
irisSummary <- lapply(irisSummary, do.call, what=rbind)
```

This produces a list of matrices with summary statistics by iris species:

```{r iris3}
irisSummary
```

Each of these can again be displayed as a nicely formatted table using `kable`
but unfortunately information about the column that was summarised will be lost
in the process. For some output formats `kable` supports the use of a `caption`
option but unfortuantely this doesn't work when producing Markdown, as is the
case here. An alternative is to use the `pander` package to produce output
suitable for further processing with pandoc.

```{r iris4, results="asis"}
suppressPackageStartupMessages(library(pander))
panderOptions('table.split.table', Inf) ## don't split tables
pander(irisSummary[1:2])
```

Alternatively the following code produces somewhat more elegant output at
the expense of a few extra lines of code.

```{r iris5, results="asis"}
for(i in 3:4){
    set.caption(sub(".", " ", names(irisSummary)[i], fixed=TRUE))
    pander(irisSummary[[i]])
}
```

The functionality provided by `pander` is alot more poweful than the simple `kable`
function and can handle a wide variety of R objects.

```{r iris6, results="asis"}
pander(t.test(Sepal.Length ~ Species=="setosa", data=iris))
```


# Converting from Markdown to multiple output formats using `knitr`
Once R code chunks have been executed via `knitr` the resulting Markdown document can be
converted to a variety of other formats with the help of `pandoc`. This generally
works well but can require the construction of lengthy command lines. To make things
worse these command lines may differ depending on the desired output format. If more
than one output format is desired this can quickly become tedious. Fortunately `knitr`
includes a function `pandoc` that takes care of the conversion process and
can use a configuration file that lists all the desired options for the desired target
formats. The configuration file used for this document is shown below^[This also
demonstrates another feature of `knitr`: It is possible to 
[include external documents](http://yihui.name/knitr/demo/child/) using the `child`
chunk option].

````
```{r configFile, child="example.pandoc"}
```
````

This file contains one block with format specific options for each output format, 
starting with `t: <format>`. Note that the first block has no target format specification
and contains options that apply to all output formats. The use of a configuration file 
like this makes it easy to manage the (potentially large) number of options required to 
achieve the desired output.  

# Preparing manuscripts for publication
Once an analysis has been completed and documented using techniques like the ones
described above it may be desirable to use it as part of a publication without having
to re-write it all. The purpose of this chapter is to investigate how well the authoring
of scientific papers in Markdown is supported by the combination of `knitr` and `pandoc`
and to demonstrate customisations to the default set-up where rquired.

## Requirements
To be able to produce manuscripts that are suitable for submission to scientific journal
several features are required. The purpose of this chapter is to explore to what extend
the combination of `knitr` and `pandoc` can deliver a publication ready manuscripts and
discuss simple extensions to add or enhance required features. 

Features essential to for a manuscript intended for submission to a journal are

* References need to be cited throughout the text and listed at the end in a format
  specified by the journal.
* Figures and tables need to be numbered and cross-references to these should be
  generated automatically, i.e. the numbers referred to in the text are updated 
  automatically if the order of figures or tables changes.
* Equations need to be rendered appropriately, numbered and cross-referenced where
  required.
* A list of author names and affiliations needs to be displayed as part of the title block.
* It has to be possible to preceed the main text with an abstract that may have to be
  formatted differently from the body of the manuscript.
* Support for footnotes.

## Document meta information
Documents may contain metadata blocks in [YAML](http://www.yaml.org/) format. These
blocks begin with three dashes `---` and end with either three dashes `---` or thee
dots `...`. More than one metadata block can be present in the same document in which
case conflicts caused by duplicate fields will be resolved by retaining the field that
occurred first.

Below is the metadata block used for this document.

````yaml
```{r child="metadata.yaml"}
```
````

Information gathered from metadata blocks is used by `pandoc` to populate metadata
fields in the output document. This can be used to set the title, list of authors
and abstract. Entries may contain (nested) lists and objects but note that the default
templates make assumptions about the structure of specific fields. The author field in 
particular is expected to be a simple list or string. For the purpose of preparing 
reports or publications it may be convenient to use a richer structure, e.g. a list 
containing objects for name, affiliation and contact details. See the chapter on 
[custom templates](#custom-templates) for details on how this structured author 
information might be used. 

## Adding a bibliography
Fortunately adding a list of refernces as well as citing them throughout the document
is well supported by `pandoc`. References need to be contained in a bibliography
file, which can be in a variety of formats (check the 
[pandoc documentation](http://johnmacfarlane.net/pandoc/README.html#citations) for
a list of supported formats). This file needs to be listed in the `biblography` 
entry of the document's meta information. A bibliography consisting of all 
references that have been cited throughout the document will be generated by `pandoc` and
added to the end of the document. The bibliography is formatted according to a 
format specified in a [CSL](http://citationstyles.org/) style file. A browsable 
repository with a large number of different styles is available at 
[http://zotero.org/styles](http://zotero.org/styles).

A citation is inserted into the text by adding the corresponding key (consisting of a '@'
followed by the citation's identifier from the database) within square brackets.
For example, `[@smith04]` would add a citation to the article with ID `smith04` to the text
and ensure that the corresponding bibliographic information is listed in the bibliography.
Several variations of this are supported by `pandoc`, see the 
[documentation](http://johnmacfarlane.net/pandoc/README.html#citations) for details.


## Better figure and table captions
We already discussed how to generate figures and tables in `knitr` and we have seen
that it is easy to add captions to these. However, so far all the figure and table cations
were plain captions without any numbering. What we would like are figure captions that
start with "Figure", or "Fig.", folowed by a number and a colon.  There currently is no
`pandoc` mechanism that allows to generate such captions in multiple output formats but
there is an active and ongoing discussion that may lead to support for this in a future
`pandoc` version. In the meantime we can use R to generate suitable labels when processing
the input document with `knitr`.

The following R function allows us to keep track of figures throughout the document,
create appropriately numbered captions as well as cross-references:

```{r fig_cap, number=TRUE}
figRef <- local({
  tag <- numeric()
  created <- logical()
  used <- logical()
  function(label, caption, prefix=options("figcap.prefix"), sep=options("figcap.sep"),
      prefix.highlight=options("figcap.prefix.highlight")) {
    i <- which(names(tag) == label)
    if(length(i) == 0){
      i <- length(tag) + 1
      tag <<- c(tag, i)
      names(tag)[length(tag)] <<- label
      used <<- c(used, FALSE)
      names(used)[length(used)] <<- label
      created <<- c(created, FALSE)
      names(created)[length(created)] <<- label
    }
    if(!missing(caption)){
      created[label] <<- TRUE
      paste0(prefix.highlight, prefix, " ", i, sep, prefix.highlight, " ", caption)
    } else {
      used[label] <<- TRUE
      paste(prefix, tag[label])
    }
  }
})
``` 

To get properly numbered figure captions all arguments to `knitr`'s `fig.cap` chunk
option have to be wrapped in a call to `figRef` with two arguments. The first argument
is the label that should be used to refer to the figure and the second argument is the 
actual figure caption. The function is designed to allow some customisation. The prefix,
e.g. 'Figure' or 'Fig.', can be set via the `prefix` argument and the separator to be used
between the number and the caption is set by the `sep` argument. It is also possible
to adjust the formatting of the prefix in the figure caption, e.g. to display it with 
strong emphasis. For convenience the desired values can be stored together with other 
R options.

Here we are setting the defaults to produce captions of the form 
"**Figure N:** caption text".

```{r figOptions}
options(figcap.prefix="Figure", figcap.sep=":", figcap.prefix.highlight="**")
```

Calling this function with the label as its sole argument will create a reference while
a call with two arguments (label and caption text) will create the actual figure caption.
Consider the following example:

```{r carDataPlot, verbatim=TRUE, fig.cap=figRef("carData", "Car speed and stopping distances from the 1920s.")}
plot(cars, xlab = "Speed (mph)", ylab = "Stopping distance (ft)",
     las = 1)
lines(lowess(cars$speed, cars$dist, f = 2/3, iter = 3), col = "red")
```

Now it is possible to refer back to this figure in the text using 
\``r rinline("figRef(\"carData\")")`\`: `r figRef("carData")` shows a plot of car speeds
and corresponding stop distances measured in the 1920s. Note the apparent 
non-linearity in the data. The log-scale data shown in `r figRef("carLogData")`
has a more linear appearance.

```{r carLogDataPlot, fig.cap=figRef("carLogData", "Car speed and stopping distances on logarithmic scales.")}
plot(cars, xlab = "Speed (mph)", ylab = "Stopping distance (ft)",
     las = 1, log = "xy")
lines(lowess(cars$speed, cars$dist, f = 2/3, iter = 3), col = "red")
```

Note how this allows to refer to figures before they are are created. Although forward 
references should generally be avoided this isn't always possible when it comes to figures.
To ensure that all figures mentioned in the text actually exist the following code
can be added to a `knitr` chunk at the end of the document.

```{r finaliser, eval=FALSE, echo=TRUE}
``` 

`r figRef("missingFigure")` dosn't exist and this generates a warning at the end 
of the document.

The same approach can be used to obtain numbered table captions and corresponding 
references in the text.

```{r tab_cap, number=TRUE}
tabRef <- local({
  tag <- numeric()
  created <- logical()
  used <- logical()
  function(label, caption, prefix=options("tabcap.prefix"), sep=options("tabcap.sep"),
      prefix.highlight=options("tabcap.prefix.highlight")) {
    i <- which(names(tag) == label)
    if(length(i) == 0){
      i <- length(tag) + 1
      tag <<- c(tag, i)
      names(tag)[length(tag)] <<- label
      used <<- c(used, FALSE)
      names(used)[length(used)] <<- label
      created <<- c(created, FALSE)
      names(created)[length(created)] <<- label
    }
    if(!missing(caption)){
      created[label] <<- TRUE
      paste0(prefix.highlight, prefix, " ", i, sep, prefix.highlight, " ", caption)
    } else {
      used[label] <<- TRUE
      paste(prefix, tag[label])
    }
  }
})

options(tabcap.prefix="Table", tabcap.sep=":", tabcap.prefix.highlight="**")
```

This can then be combined with the `pander` table generation technique demonstrated
above.

```{r carFitPlot, fig.cap=figRef("carFit", "Polynomial regression fits."), results="asis"}
plot(cars, xlab = "Speed (mph)", ylab = "Stopping distance (ft)",
    las = 1, xlim = c(0, 25))
d <- seq(0, 25, length.out = 200)
for(degree in 1:4) {
  fm <- lm(dist ~ poly(speed, degree), data = cars)
  assign(paste("cars", degree, sep = "."), fm)
  lines(d, predict(fm, data.frame(speed = d)), col = degree)
}
legend("topleft", legend=1:4, col=1:4, lty=1)

set.caption(tabRef("carFit", "ANOVA table for polynomial regression fits to car speed and stopping distance data"))
pander(anova(cars.1, cars.2, cars.3, cars.4))

```

This approach to figure and table captions has the advantage that it works for any
output format and as such is well suited for situations where multiple output formats
are required. The downside is that it doesn't make use of any cross-referencing facilities
that may be supported by one or several of the output formnats. For example, LaTeX has
excellent support for this already and HTML output would benefit from the use of links 
to the actual figure or table. When a single output format is used it clearly makes 
sense to utilise the features it provides as much as possible. The multi-format 
approach presented here could be improved through the addition of some additional
markup and a [pandoc filter](http://johnmacfarlane.net/pandoc/scripting.html) that
turns this markup into format specific output.

## Structured author information
By default `pandoc` only supports simple strings (or a list of strings) for the 
author field in the metadata block. This means that including information in
addition to author names, e.g. affiliations and addresses, is dificult (but note
that strings are interpreted as markdown so some formatting is possible). To really
support the generation of publication ready documents the use of more structured
author fields is desirable. For this document we use author information of the 
following form:

```.yaml
author: 
  - name: Author Name
    affiliation: 1
address:
  - code: 1
    address: Department, Institution, Street address
``` 
Other fields could be added to this, e.g. to indicate corresponding authors. In
order for this additional information to be displayed in the output we need to 
extend the default templates to use this information. The required changes
to the the HTML and LaTeX templates are discussed [below](#structured-author).

## Equations
There is good support for formula rendering in `pandoc`. Formulas can be written
as TeX formulas between `$` (for inline math) or `$$` (for display math). These
will be rendered in an output format specific way. In some formats, like HTML, 
the result depends on the command-line options used. 

----

```markdown
This is a familiar bit of inline math: $E=mc^2$.
```

This is a familiar bit of inline math: $E=mc^2$.

----

----

```markdown
Here is an equation in display math mode:
$$f(x, \mu, \sigma) = \frac{1}{\sigma\sqrt{2\pi}}e^{-\frac{(x-\mu)^2}{2\sigma^2}}$$
```

Here is an equation in display math mode:
$$f(x, \mu, \sigma) = \frac{1}{\sigma\sqrt{2\pi}}\,e^{-\frac{(x-\mu)^2}{2\sigma^2}}$$

----

While this generally works fairly well it doesn't allow for numbered equations.
A possible workaround is to use `pandoc`'s example list feature for this purpose.
An axample list consists of consecutively numbered list elements that don't have to
be placed within the same list, i.e. they can be placed throughout the document.

----

```markdown
(@cauchy) $f(x) = \frac{1}{\pi(1+x^2)}$

The Cauchy distribution (with density given in Eq. (@cauchy)) is a special case 
of Student's $t$-distribution (Eq. (@tdist)) with $\nu = 1$.

(@tdist) $$f(t; \nu) = \frac{\Gamma(\frac{\nu+1}{2})} {\sqrt{\nu\pi}\,\Gamma(\frac{\nu}{2})} \left(1+\frac{t^2}{\nu} \right)^{-\frac{\nu+1}{2}}$$
```

(@cauchy) $f(x) = \frac{1}{\pi(1+x^2)}$

The Cauchy distribution (with density given in Eq. (@cauchy)) is a special case 
of Student's $t$-distribution (Eq. (@tdist)) with $\nu = 1$.

(@tdist) $$f(t; \nu) = \frac{\Gamma(\frac{\nu+1}{2})} {\sqrt{\nu\pi}\,\Gamma(\frac{\nu}{2})}\,\left(1+\frac{t^2}{\nu} \right)^{-\frac{\nu+1}{2}}$$

----

While this solves the basic problem of getting numbered equations it isn't perfect. Equations
are not centred and numberes appear on the left rather than the right as is customary.
An additional problem when using display math (as in Eq. (@tdist)) is that the number
and the equation are not lined up properly. It is possible to fix this in HTML output
through the use of appropriate CSS but that doesn't help for other output formats. 

----

```markdown
<div class="equation">
(@gamma) $$\Gamma(t) = \int^\infty_0 x^{t-1}e^{-x}dx$$
</div>
```

<div id="gamma" class="equation_css">
(@gamma) $$\Gamma(t) = \int^\infty_0 x^{t-1}e^{-x}dx$$
</div>

----

In the above example the equation is wrapped in a `div` with class "equation". This allows
application of suitable CSS to improve the alignment of the equation. 
Horizontal alignment of the formula is relatively straightford with CSS so it can be
centred without too much difficulty in HTML output. Proper alignment with the automatically 
generated number proves to be more difficult. The following javascript code does the trick.

````javascript
```{r child="include/equation.js"}
```  
````

This isn't particularly elegant and only solves the problem for HTML.
For LaTeX the lack of proper equation handling is particularly unsatisfying as
LaTeX  has much better support for equations. Again using a 
[filter](http://johnmacfarlane.net/pandoc/scripting.html) for additional 
processing to produce equations that are better suited to the output format. This should
allow the use of LaTeX equation environments in LaTeX output and could be used
to produce better HTML output as well. This filter can also make use of the `div` 
introduced above. See [below](#pandoc-filters) for an example of how this can be achieved.

## Footnotes
The use of footnotes is well supported by `pandoc`. The easiest way to add a footnote is 
the inline syntax.

----

```markdown
This is regular text^[with a footnote].
```

This is regular text^[with a footnote].

----

It is also possible to use labels to identify a footnote, similar to the way references work.

----

````markdown
When using the reference style a short label is present in the text[^1] and the
actual footnote text is defined elsewhere.

[^1]: This is closer to the appearance of the rendered text in the output but updating
the footnote text is a little bit more work since it may be somewhere else in the document.

    The advantage of this format is that it supports multi-block content. It could even
    contain a code block if desired.
    
    ```r
    message("This is R code")
    ```
    
    The trick is to indent subsequent paragraphs to indicate that they form part of the
    footnote.
    
Afterwards the normal text continues.
````

When using the reference style a short label is present in the text[^1] and the
actual footnote text is defined elsewhere.

[^1]: This is closer to the appearance of the rendered text in the output but updating
the footnote text is a little bit more work since it may be somewhere else in the document.

    The advantage of this format is that it supports multi-block content. It could even
    contain a code block if desired.
    
    {code.block}
    
    The trick is to indent subsequent paragraphs to indicate that they form part of the
    footnote.
    
Afterwards the normal text continues.

----

# Customising output
`Pandoc` provides default templates for all supported output formats. These
provide a quick and easy way to generate output in a variety of formats.
Typically they produce decent looking results and in some cases will be
all that is required. However, complex documents for large projects often
benefit from some customisation. 

## Using custom headers
Many output formats use header information to specify details of the 
output rendering. Modifying the header can give substantial control 
over the appearance of the final output. Using the `-H` option of 
`pandoc` allows the contents of arbitrary files to be added to the
header of a template. For example, to tweak the appearance of figure 
and table captions the following file is included for the LaTeX output
of this document.

````latex
```{r child="include/captions.tex"}
```
````
Multiple files with header content can be included in this way, allowing
for a some flexibility and the option to create re-usable header fragments.

## Custom templates
Sometimes more substantials modifications are called for. In this case
it may be possible to create a custom template that incorporates the 
desired changes. The best way to create a custom template is to start
with the default template for the desired output format and modify it as
necessary. The latest version of all `pandoc` templates are available
on [GitHub](https://github.com/jgm/pandoc-templates). While it is possible
to simply download the desired file and modify it, the recommended way
of producing custom templates is to fork the repository and then
commit all changes to the forked version. This makes it easier to update
the customised version of the template with changes to the 
default^[see <https://help.github.com/articles/syncing-a-fork> for a description of
how to do this]. This may be necessary to accomodate changes to `pandoc`.

Templates allow access to `pandoc`'s template variables. All variable expressions
are surrounded by `$`: `$variable$`. These can be used
in conditionals, adjusting the contents of the output accordingly.

```
$if(variable)$
$variable$
$else$
Some default text.
$endif$
```
In cases where a variable contains a list it is possible to iterate over
its elements, inserting each into the text. For exxample, the default latex 
template has the following code to populate the author field.

```
$if(author)$
\author{$for(author)$$author$$sep$ \and $endfor$}
$endif$
```
Note that `$sep$` allows the specification of a separator to be used between
list elements.

### Including an abstract in HTML output
Customising templates can be particularly useful when the default doesn't
handle certain types of metadata that should be displayed in the output.
For example, this document contains an abstract that is included in the 
PDF output but ignored in the HTML output by default. This can be changed by
adding a few lines to the HTML5 template.

```
$if(abstract)$
<section class="abstract">
<h1 class="abstract">Abstract</h1>
$abstract$
</section>
$endif$
```
Adding this between the header block and the table of contents block has the 
desired effect of including the text of the abstract near the top of the
page. This can then be styled with CSS as desired. The style sheet used here
has the following:

```css
h1.abstract
	{
	text-align: center;
	}
section.abstract
	{
	width:50%;
	margin-left:auto;
	margin-right:auto;
	background:#eef;
	padding:10px;
	}
```

### Using structured author fields {#structured-author}
LaTeX, unsurprisingly, has pretty good support for additional author information.
Here we use the [`authblk`](http://www.ctan.org/pkg/authblk) package to typeset 
the addresses. This allows us to specify authors and their affiliations
through a shared identifier. This conveniently matches the format used
in the metadata block. 

Here is the (somewhat more complex) template code to insert author
information with optional affiliation.
```{.latex .numberLines}
$if(author)$
	\usepackage{authblk}
	$if(address)$
		$for(author)$
			\author[$author.affiliation$]{$author.name$}
		$endfor$
		$for(address)$
			\affil[$address.code$]{$address.address$}
		$endfor$
	$else$
		$for(author)$
			$if(author.name)$
				\author{$author.name$}
			$else$
				\author{$author$}
			$endif$
		$endfor$ 
	$endif$
$endif$
```

This supports multiple authors and can handle both the new structured
author blocks as well as the original plain author field. However,
in its current for it only supports a single affiliation per author.
Note how line 5 uses the `author.affiliation` field:
`\author[$author.affiliation$]{$author.name$}`. This assumes a single
string but can be modified to handle a list of strings instead by
wrapping the variable in a for loop: 
`\author[$for(author.affiliation)$$author.affiliation$$sep$, $endfor$]{$author.name$}`.
This largely works as intended with the only limitation that the
IDs used to associate authors with intitutions are directly used
as superscripts to the author's name. It would be desirable to generate
a sequence of consecutive numbers (or other LaTeX symbols) automatically
but that would require further processing.

As one might expect direct support for structured author information
in HTML output is a bit more limited. The modification to the template
follow similar lines.

```{.html .numberLines}
$if(author)$
	<h2 class="author">
	$for(author)$
		$if(author.name)$
			$if(author.affiliation)$
				<span data-affiliation="$for(author.affiliation)$$author.affiliation$$sep$, $endfor$" class="author_name">$author.name$</span>
			$else$
				$author.name$
			$endif$
		$else$
			$author$
		$endif$$sep$, 
	$endfor$
	</h2>
	$if(address)$
		<div class="address">
		<a href="#" id="address-header" onclick=toggleDisplay("address-list")>Author Affiliations</a>
		<ol id="address-list" style="display:none">
		$for(address)$
			<li data-id="$address.code$">$address.address$</li>
		$endfor$
		</ol>
		</div>
	$endif$
$endif$
```
The addresses are added as an ordered list (lines 17 - 25), providing us with 
automatic numbering. The corresponding IDs are stored in a data attribute
of the list elements as well as with the author names. To generate the matching
superscripts to the names we use a few lines of javascript.

````{.javascript .numberLines}
```{r affiliationCode, child="include/affiliation.js"}
```
````
Unlike the LaTeX solution this supports the use of arbitrary IDs but 
requires some extra processing of the output. It also relies on javascript
being enabled in the browser. The appearance of the address information can then 
be modified with CSS. Here we also use additional javascript to toggle
the display when the heading is clicked.

# Pandoc filters
It is possible to add additional processing steps to format conversions
in `pandoc` through the use of filters. A filter is a small (or possibly
rather complex) program that is executed by `pandoc` after the contents of
the input file has been transformed into `pandoc`'s native (JSON) 
representation. The filter can then modify that representation as desired
and the resulting document is then converted to the target format.
The filter has access to the requested target format and can therefore 
be used to make format specific modifications. This can, e.g., be used
to wrap equations in appropriate environments in LaTeX.

Since `pandoc` is written in [Haskell](http://www.haskell.org/haskellwiki/Haskell)
this is also the language best suited to writing filters. However, there
is a Python module ([pandocfilters](https://github.com/jgm/pandocfilters)) providing
support for the parsing and writing of `pandoc`'s native format. Once a filter
has been written, and assuming that it is available in an executable form,
it can be passed to `pandoc` through the `--filter` command-line option. 


## Better equations 
For the remainder of this chapter we will discuss a filter designed to provide
better equations in LaTeX and HTML output. The aim is to use the `equation`
environment in LaTeX for individual, numbered equations and the `align` environment
for groups of equations that should be lined up with each other. In HTML output
these should be rendered in a similar way, i.e. equations are centred with 
numbering on the right and equations are lined up correctly where applicable.

Here are a few equations we will use for testing:

----

```markdown
<div id="volterra" class="equation">
(@volterra) $$\frac{dx}{dt} = x(\alpha - \beta y)$$
$$\frac{dy}{dt} = - y(\gamma - \delta  x)$$
</div>

The system of ordinary differential equations given by 
Eq. <span id="volterra" class="eq_ref">(@volterra)</span> 
is commonly used to describe predator-prey systems. Any solution to this
system of equations satisfies the equality in 
Eq. <span id="volterra_constant" class="eq_ref">(@volterra_constant)</span>.

<div id="volterra_constant" class="equation">
(@volterra_constant) $$V = -\delta \, x + \gamma \, \log(x) - \beta \, y + \alpha \, \log(y)$$
</div>

```

<div id="volterra" class="equation">
(@volterra) $$\frac{dx}{dt} = x(\alpha - \beta y)$$
$$\frac{dy}{dt} = - y(\gamma - \delta  x)$$
</div>

The system of ordinary differential equations given by 
Eq. <span id="volterra" class="eq_ref">(@volterra)</span> 
is commonly used to describe predator-prey systems. Any solution to this
system of equations satisfies the equality in 
Eq. <span id="volterra_constant" class="eq_ref">(@volterra_constant)</span>.

<div id="volterra_constant" class="equation">
(@volterra_constant) $$V = -\delta \, x + \gamma \, \log(x) - \beta \, y + \alpha \, \log(y)$$
</div>

----

### Python filter implementation
This section provides a step-by-step explanation of the filter implementation. 
The full code of the final script is available in the [appendix](#equation-filter).
The Python script for achieving the desired equation formatting needs to
import the `pandocfilters` module. We will also need regular expressions.

```python
#! /usr/bin/env python

from pandocfilters import toJSONFilter, RawBlock, Div
import re
```

The `pandocfilters` module provides the function `toJSONFilter`. This function
takes another function as its sole argument and applies it to each node of
the JSON document that is read from standard input.  To be able to pass this
script to `pandoc` as one of the command-line options it needs to include the 
following two lines.

```python
if __name__ == '__main__':
    toJSONFilter(equation)
```

Of course we also need to define the function `equation`. This function has
to accept four arguments corresponding to the key and value of the current
node, the requested target format and the document's meta information.

```python
def equation(key, value, format, meta):
  # process equation nodes
```
Whenever this function returns a value this value will replace the current node.
If nothing is returned the current node is unchanged. The return value can be a 
single object or a list of objects. To this end we will need to 
generate Objects representing raw LaTeX and HTML code to add the desired formatting.
To this end we define the following two functions with the help of element 
constructors provided by `pandocfilters`.

```python
def latex(x):
  return RawBlock('latex',x)

def html(x):
    return RawBlock('html', x)
``` 

The key is a short string identifying the type of node that is currently being
processed. Since we have chosen to wrap all equations in a `div` with class
'equation' we need to check this and proceed with processing whenever one 
of these `div`s is encountered.

```python
  if key == 'Div':
    [[ident,classes,kvs], contents] = value
    if 'equation' in classes:
      # process equation
```

Once we have identified the `div` we need to extract the actaul equations from the
`contents`. Note that this may contain either one or several equations and these
may be wrapped in an ordered list (if the example list style numbering system is
used for other output formats). The following function extracts the contents of 
math environments.

```python
def getMath(x):
    if isinstance(x, list):
        return [getMath(l) for l in x]
    if isinstance(x, dict):
        if x['t'] == 'Math':
            return x['c'][1]
        else:
            return getMath(x['c'])
```

We need to apply this to all sub-nodes of the div. To facilitate the traversal of
the corresponding nested list the function below is used.

```python
def iter_flatten(iterable):
  it = iter(iterable)
  for e in it:
    if isinstance(e, (list, tuple)):
      for f in iter_flatten(e):
        yield f
    else:
      yield e
```

Together these functions allow us to generate a list of equations:

```python
math = iter_flatten([ getMath(contents)])
math = [x for x in math if x is not None]
```

Note that this produces a list of lists with elements that may be 
empty or `None` (because not all nodes contain equations). 
The list comprehension in the second line above is used to flatten the list
and remove all unwanted entries.

###LaTeX output
With the actual equations extracted from the div the main task remaining
is to format them appropriately for the desired output format. If the div
contains a single equation we'll use the `equation` environment for
LaTeX output. If there are multiple equations grouped to gether the `align`
environment is used to create a nicely lined-up block of equations. The `align`
environment in LaTeX uses `&` to define alignment points. We want to allow manual alignment 
of equations, meaning that existing `&` symbols need to be respected. In the absence
of a predefined alignment we'll align equations on the first relational operator^[The 
`align` environment allows the use of multiple alignment points to group equations 
into columns. Automatic alignment generated by this filter
only supports a single alignment point, set at the first relational operator. 
If you want a more complex alignment all equations in a block should define the
required alignment points.]. This is achieved with the function below.

```python
def alignLatexMath(x):
    global relSymbol
    relPattern = '|'.join(relSymbol)
    if re.search(r'[^\\]&', x) is None:
         idx = re.search(relPattern, x).start()
         return x[:idx] + '&' + x[idx:]
    return x
```  

The following code fragment then generates the LaTeX output.

```python
if format == 'latex':
    type = 'equation'
    if len(math) > 1:
        type = 'align'
        math = [alignLatexMath(x) for x in math]
    if ident != '':
        label = '\\label{' + ident + '}'
    else:
        type = type + '*'
    return [latex('\\begin{' + type + '}' + label + "\n" + \
                  "\\\\\n".join(math) + \
                  "\n" + '\\end{' + type + '}')]
```

Note that this adds a label to the environment based on the ID of the div (if present)
and for equations without label the numbering is suppressed. In cases where the
`align` environment is used only the first equation will be labelled with the provided
ID and all subsequent equations will be assigned labels of the form `ID.n`, starting
with n = 1 and increasing by one for each subsequent equation. While this allows 
referencing of individual equations within an `align` environment it can be difficult
to maintain these references if new equations get inserted into the middle of a 
block.

### HTML output
For HTML output we'll use a table with appropriate CSS styling to obtain a similar
effect. Since we can't rely on LaTeX's build-in facilities for handeling equations
this requires slightly more effort. 

The output consists of three columns per equation. These are used for the left and 
right sides of the equation with the relational operator in the middle. If multiple
columns of equations are required these three columns will be repeated for each.
If the `div` is labelled an additional column is added at the end to contain the
equation number^[Here we are using a global variable (`eqCount`) in the python script to 
keep track of the number of equations. We could use a CSS counter instead to insert
the number into the web page as it is rendered in the browser. If all we wanted to do
was to number the equations that would probabaly be the better solution but we will
proceed to extend this to allow cross-references as well and CSS can't really handle
those.]. 

To prepare equations for formatting we split them at the alignment point.

```python
def alignHtmlMath(x):
    global relSymbol
    relPattern = r'|'.join(relSymbol)
    cols = re.split(r"[^\\]&(?!" + relPattern + ")", x)
    align = [re.search(r"[^\\]&", x) for x in cols]
    out = []
    for i in range(len(align)): 
        if align[i] is not None:
            idx = align[i].start()
            skip = 1
        else:
            idx = re.search(relPattern, cols[i]).start()
            skip = 0
        out = out + [[cols[i][:idx+skip], cols[i][idx+2*skip:idx+2*skip+1], 
                     cols[i][idx+2*skip+1:]]] 
    return out
    
math = [alignHtmlMath(x) for x in math]
```

Now we only need to decorate the resulting equation fragments with appropriate
HTML tags to create the table.
 
```python
if ident != '':
    label = 'id=\"' + ident + '\" '
head = [html('<table ' + label + 'class=\"' + ' '.join(classes) + '\" ' + \
    ' '.join(kvs) + '>' + "\n")]
tail = [html('</table>' + "\n")]
body = [html('<tbody>' + "\n")]
for eq in math:
    eqCount = eqCount + 1
    body = body + [html('<tr>' + "\n")] 
    for sub in [formatHtmlMath(y) for y in eq]:
        body = body + sub
    if ident != '':
        body = body + [html(' <td class=\"eq_number\"> <br>(' + \
                            str(eqCount) + ')<br> </td>')]
    body = body + [html('</tr>' + "\n")]
body = body + [html('</tbody>' + "\n")]
```

### Styling equations with CSS
Now that we have a filter that generates HTML output that is structured 
to allow proper layout of equations the only missing piece is the actual 
layout. A little bit of CSS should take care of that.

We start by setting the table to span the full width of the content `div`.
This will allow us to centre the equation while displaying the equation 
number at the right margin.

```css
table.equation 
{
	width: 100%;
}
```

Unfortunately this results in spreading the equation out across the entire
line. To achieve the desired effect we define a `nostretch` style for table cells
and apply it to the cells holding the central and right part of the equation. 

```css
td.nostretch
{
	width: 1%;
}
```

The result of this is that the the left and rightmost cells will stretch to fit
the width of the contents `div`. All that is left to do is to ensure that the
cotents of these cells is aligned properly.

```css
table.equation td:not(nostretch)
{
	text-align: right;
}

table.equation td.eq_right
{
	text-align: left;
}
```

We need an extra rule to ensure that the right part of the equation is aligned left
so that it is flush with the central part even when there are multiple rows
with different length of contents for this column.

### Keeping track of references
Now equations are properly displayed and numbered in both HTML and LaTeX 
output but unfortunately the cross-references in the text (generated via the 
example list mechanism) are no longer garuanteed to match the numbers of the 
actual equations. In HTML output the numbers should match if all equations
contained in `equation` `div`s are numbered but LaTeX may use a different
numbering scheme, e.g. numbering equations by chapter. Again it would make sense
to utilise LaTeX's inherent abilities (this time for cross-referencing) and
to add some additional code to achieve the same effect with HTML. It would 
also be nice to retain the example list method for numbering equations
as a fall back for other output formats.

To achieve this we will again resort to additional markup, in this case 
a `span` element: `markdown <span id="label" class="eq_ref">(@label)</span>`.
This can the be processed by our equation filter by adding just a few extra
lines of code.

```python
if key == 'Span':
	[[ident,classes,kvs], contents] = value
	if 'eq_ref' in classes:
	    if format == 'latex':
	        return latexInline("(" + "\\ref{" + ident + "})")              
```

For convenience we have also introduced a new (very simple) function.

```python
def latexInline(x):
    return RawInline('latex', x)
```

For HTML output things are a bit more complicated as we have to keep track
of equation numbers. We'll use a dictionary to store labels and corresponding 
numbers for equations. Whenever a new equation label is encountered in the
document it has to be added to the dictionary. The function `eqNumber` can be
called with a label and will return the corresponding equation number (registering
the label in the process if it hasn't occoured before).

```python
_eqLabel = {}
def eqNumber(id):
    global _eqLabel
    if id not in _eqLabel.keys():
        _eqLabel[id] = len(_eqLabel) + 1
    return str(_eqLabel[id])
```




# Appendix {-}
## Equation filter
````{.python .numberLines}
```{r equation-filter-complete, child="include/equation.py"}
```
````

## Session info
```{r session-info, echo=FALSE, results="asis"}
sessionInfo()
```

```{r finaliser, eval=TRUE, echo=FALSE}
if(!all(environment(figRef)$created)){
    missingFig <- which(!environment(figRef)$created)
    warning("Figure(s) ", paste(missingFig, sep=", "), " with label(s) '", 
      paste(names(environment(figRef)$created)[missingFig], sep="', '"),
      "' are referenced in the text but have never been created.")
}
if(!all(environment(figRef)$used)){
    missingRef <- which(!environment(figRef)$used)
    warning("Figure(s) ", paste(missingRef, sep=", "), " with label(s) '", 
      paste(names(environment(figRef)$used)[missingRef], sep="', '"),
      "' are present in the document but are never referred to in the text.")
}
```
